---
title: "Statistical Arbitrage Strategy"
author: "Jacques Joubert"
output: word_document
---



##Import packages and set directory
The first step is always to import the packages needed.

```{r , message=FALSE, warning=FALSE}
#Imports
require(tseries)
require(urca) #Used for the ADF Test
require(PerformanceAnalytics)
```

This strategy will be run on shares listed on the Johannesburg Stock Exchange (JSE); because of this I wont be using the quantmod package to pull data from yahoo finance, instead I have already gotten and cleaned the data that I stored in a SQL database and moved to csv files on the Desktop.

I added all the pairs used in the strategy to a folder which I now set to be the working directory.

```{r}
##Change this to match where you stored the csv files folder name FullList
setwd("C:\\Users\\Administrator\\Documents\\GitHub\\QuantInsti-Final-Project-Statistical-Arbitrage\\database\\FullList")
```

##Functions that will be called from within other functions (No user interaction)
Next: Create all the functions that will be needed. The functions below will be called from within other functions so you dont need to worry about the arguments.

###AddColumns
The AddColumns function is used to add columns to the dataframe that will be needed to store variables. 

```{r}
#Add Columns to csvDataframe
AddColumns <- function(csvData){
  csvData$spread <- 0
  csvData$adfTest <- 0
  csvData$mean <- 0
  csvData$stdev <- 0
  csvData$zScore <- 0
  csvData$signal <- 0
  csvData$BuyPrice <- 0
  csvData$SellPrice <- 0
  csvData$LongReturn <- 0
  csvData$ShortReturn <- 0
  csvData$Slippage <- 0
  csvData$TotalReturn <- 0
  return(csvData)
}
```
###PrepareData
The PrepareData function calculates the pair ratio and the log10 prices of the pair. It also calls the AddColumns funtion within it.

```{r}
PrepareData <- function(csvData){
  #Calculate the Pair Ratio
  csvData$pairRatio  <-  csvData[,2] / csvData[,3]
  
  #Calculate the log prices of the two time series
  csvData$LogA <- log10(csvData[,2])
  csvData$LogB <- log10(csvData[,3])
  
  #Add columns to the DF
  csvData <- AddColumns(csvData)
  
  #Make sure that the date column is not read in as a vector of characters
  csvData$Date <- as.Date(csvData$Date)
  
  return(csvData)
}
```
###GenerateRowValue
The GenerateRowValue function Calculates the mean, standard deviation and the z-score for a given row in the dataframe.

```{r}
#Calculate mean, stdDev, and z-score for the given Row [end]
GenerateRowValue <- function(begin, end, csvData){
  average <- mean(csvData$spread[begin:end])
  stdev <- sd(csvData$spread[begin:end])
  
  csvData$mean[end] <-  average
  csvData$stdev[end] <- stdev
  csvData$zScore[end] <- (csvData$spread[end]-average)/stdev
  
  return(csvData)  
}
```
###GenerateSignal
The GenerateSignal function creates a long, short, or close signal based on the z-score. You can manually change the z-score. I have set it to 1 and -1 for entry signals and any z-score between 0.5 and -0.5 will create a close/exit signal.

```{r}
#Generate trading signals based on a z-score of 1 and -1 
GenerateSignal <- function(counter, csvData){
  #Trigger and close represent the entry and exit zones (value refers to the z-score value)
  trigger  <- 1
  close  <-  0.5
  
  currentSignal <- csvData$signal[counter]
  prevSignal <- csvData$signal[counter-1]
  
  #Set trading signal for the given [end] row
  if(csvData$adfTest[counter] == 1)
  {
    #If there is a change in signal from long to short then you must allow for the 
    #current trade to first be closed
    if(currentSignal == -1 && prevSignal == 1)
      csvData$signal[counter] <- 0
    else if(currentSignal == 1 && prevSignal == -1)
      csvData$signal[counter] <- 0
    
    #Create a long / short signal if the current z-score is larger / smaller than the trigger value
    #(respectively)
    else if(csvData$zScore[counter] > trigger)
      csvData$signal[counter] <- -1
    else if (csvData$zScore[counter] < -trigger)
      csvData$signal[counter] <- 1
    
    #Close the position if z-score is beteween the two "close" values
    else if (csvData$zScore[counter] < close && csvData$zScore[counter] > -close)
      csvData$signal[counter] <- 0
    else 
      csvData$signal[counter] <- prevSignal
  }
  else 
    csvData$signal[counter] <- 0
  
  return(csvData)
}
```
###GenerateTransactions
The GenerateTransactions function is responsible for setting the entry and exit prices for the respective long and short positions needed to create a pair.

Note: QuantInsti taught us a very specific way of backtesting a trading strategy. They used excel to teach strategies and when I coded this strategy I used a large part of the excel methodology. 

Going forward however I would explore other ways of storing variabels. One of the great things about this method is that you can pull the entire dataframe and analys why a trade was made and all the details pertaining to it. 

```{r}
#Transactions based on trade signal
#Following the framework set out initially by QuantInsti (Note: this can be coded better) 
GenerateTransactions <- function(currentSignal, prevSignal, end, csvData){
  #In a pair trading strategy you need to go long one share and short the other
  #and then reverse the transaction when you close
  
  ##First Leg of the trade (Set Long position)
  #If there is no change in signal
  if(currentSignal == 0 && prevSignal == 0)
    csvData$BuyPrice[end] <- 0    
  else if(currentSignal == prevSignal)
    csvData$BuyPrice[end] <- csvData$BuyPrice[end-1]     
  
  #If the signals point to a new trade
  #Short B and Long A
  else if(currentSignal == 1 && currentSignal != prevSignal)
    csvData$BuyPrice[end] <- csvData[end, 2] 
  #Short A and Long B
  else if(currentSignal == -1 && currentSignal != prevSignal){
    csvData$BuyPrice[end] <- csvData[end, 3] * csvData$pairRatio[end]
    transactionPairRatio <<- csvData$pairRatio[end]
  }
  
  #Close trade
  else if(currentSignal == 0 && prevSignal == 1)
    csvData$BuyPrice[end] <- csvData[end, 2] 
  else if(currentSignal == 0 && prevSignal == -1)
    csvData$BuyPrice[end] <- csvData[end, 3] * transactionPairRatio  
  
  
  
  ##Second Leg of the trade (Set Short position)
  ##Set Short Prices if there is no change in signal
  if(currentSignal == 0 && prevSignal == 0)
    csvData$SellPrice[end] <- 0    
  else if(currentSignal == prevSignal)
    csvData$SellPrice[end] <- csvData$SellPrice[end-1] 
  
  #If the signals point to a new trade
  else if(currentSignal == 1 && currentSignal != prevSignal){
    csvData$SellPrice[end] <- csvData[end, 3] * csvData$pairRatio[end]
    transactionPairRatio <<- csvData$pairRatio[end]
  }
  else if(currentSignal == -1 && currentSignal != prevSignal)
    csvData$SellPrice[end] <- csvData[end, 2] 
  
  #Close trade
  else if(currentSignal == 0 && prevSignal == 1){
    csvData$SellPrice[end] <- csvData[end, 3] * transactionPairRatio
  }
  else if(currentSignal == 0 && prevSignal == -1)
    csvData$SellPrice[end] <- csvData[end, 2] 
  
  return(csvData)
}
```
###GetReturns
GetReturns calculates the returns on each position after it has been closed and then calculates the total returns and adds slippage.

```{r}
#Calculate the returns generated after each transaction
#Add implementation shortfall / slippage
GetReturns <- function(end, csvData, slippage){
  #Calculate the returns generated on each leg of the deal (the long and the short position)
  #Long leg of the trade
  if(csvData$signal[end] == 0 && csvData$signal[end-1] != 0 )
    csvData$LongReturn[end] <- (csvData$BuyPrice[end] / csvData$BuyPrice[end-1]) - 1
  #Short Leg of the trade
  if(csvData$signal[end] == 0 && csvData$signal[end-1] != 0 )
    csvData$ShortReturn[end] <- (csvData$SellPrice[end-1] / csvData$SellPrice[end]) - 1
  
  #Add slippage
  if(csvData$ShortReturn[end] != 0)
    csvData$Slippage[end] <- slippage
  
  #If a trade was closed then calculate the total return
  if(csvData$ShortReturn[end] != 0 && csvData$LongReturn[end] != 0)
    csvData$TotalReturn[end] <- ((csvData$ShortReturn[end] + csvData$LongReturn[end]) / 2) + csvData$Slippage[end]
  
  return(csvData)
}
```

###GenerateReports
The next two arguments are used to generate reports. A report includes the following: 
Charting:
1. An Equity curve
2. Drawdown curve
3. Daily returns bar chart

Statistics:
1. Annual Retruns
2. Annualized Sharpe Ratio
3. Maximum Drawdown
4. Total trades
5. Success ratio
6. PnL ratio

Table:
1. Top 5 drawdowns and their duration

Note: If you have some extra time then you can further break this function down into smaller functions inorder to reduce the lines of code and improve userbility. Less code = Less Bugs

```{r}
#Returns an equity curve, annualized return, annualized sharpe ratio, and max drawdown
GenerateReport <- function(pairData, startDate, endDate){
  #Subset the dates 
  returns  <-  xts(pairData$TotalReturn, as.Date(pairData$Date))
  returns  <-  returns[paste(startDate,endDate,sep="::")]
  
  #Plot
  charts.PerformanceSummary(returns)
  
  #Metrics
  print(paste("Annual Returns: ",Return.annualized(returns)))
  print(paste("Annualized Sharpe: " ,SharpeRatio.annualized(returns)))
  print(paste("Max Drawdown: ",maxDrawdown(returns)))
  
  #var returns = xts object
  totalTrades  <-  0
  positiveTrades  <-  0
  profitsVector  <- c()
  lossesVector  <- c()
  
  #loop through the data to find the + & - trades and total trades
  for(i in returns){
    if(i != 0){
      totalTrades  <- totalTrades + 1
      if(i > 0){
        positiveTrades  <- positiveTrades + 1
        profitsVector  <- c(profitsVector, i)
      }
      else if (i < 0){
        lossesVector  <- c(lossesVector, i)
      }
    }
  }
  
  #Print the results to the console
  print(paste("Total Trades: ", totalTrades))
  print(paste("Success Rate: ", positiveTrades/totalTrades))
  print(paste("PnL Ratio: ", mean(profitsVector)/mean(lossesVector*-1)))
  print(table.Drawdowns(returns))
  
}

#Use this one if you have the returns in xts format and want to generate a report
GenerateReport.xts <- function(returns, startDate = '2005-01-01', endDate = '2015-11-23'){
  returns  <-  returns[paste(startDate,endDate,sep="::")]
  
  #Plot
  charts.PerformanceSummary(returns)
  
  #Metrics
  print(paste("Annual Returns: ",Return.annualized(returns)))
  print(paste("Annualized Sharpe: " ,SharpeRatio.annualized(returns)))
  print(paste("Max Drawdown: ",maxDrawdown(returns)))
  
  #var returns = xts object
  totalTrades  <-  0
  positiveTrades  <-  0
  profitsVector  <- c()
  lossesVector  <- c()
  
  #Itterate through data to get the + & - trades
  for(i in returns){
    if(i != 0){
      totalTrades  <- totalTrades + 1
      if(i > 0){
        positiveTrades  <- positiveTrades + 1
        profitsVector  <- c(profitsVector, i)
      }
      else if (i < 0){
        lossesVector  <- c(lossesVector, i)
      }
    }
  }
  
  #Print results to Console
  print(paste("Total Trades: ", totalTrades))
  print(paste("Success Rate: ", positiveTrades/totalTrades))
  print(paste("PnL Ratio: ", mean(profitsVector)/mean(lossesVector*-1)))
  print(table.Drawdowns(returns))
  
}
```

##Functions that the user will pass parameters to

The next two functions are the only functions that the user should fiddle with. 

###BacktestPair
BacktestPair is used when you want to run a backtest on a trading pair (the pair is passed in via the csv file)

Functions arguments: 

* pairData = the csv file date
* mean = the number of observations used to calculate the mean of the spread.
* slippage = the amount of basis points that act as brokerage as well as slippage
* adfTest = a boolean value - if the backtest should test for co-integration
* criticalValue = Critical Value used in the ADF Test to test for co-integration
* generateReport = a boolean value - if a report must be generated
  
```{r}
#The function that will be called by the user to backtest a pair
BacktestPair <- function(pairData, mean = 35, slippage = -0.0020, adfTest = TRUE, criticalValue = -2.58, startDate = '2005-01-01', endDate = '2014-11-23', generateReport = TRUE){
  # At 150 data points
  # Critical value at 1% : -3.46
  # Critical value at 5% : -2.88
  # Critical value at 10% : -2.57
  
  #Prepare the initial dataframe by adding columns and pre calculations
  pairData <- PrepareData(pairData)
  
  #Itterate through each day in the time series
  for(i in 1:length(pairData[,2])){
    
    #For each day after the amount of days needed to run the ADF test
    if(i > 130){
      begin  <-  i - mean + 1
      end  <-  i
      
      #Calculate Spread
      spread  <-  pairData$pairRatio[end]
      pairData$spread[end]  <-  spread
      
      #ADF Test for co-integration
      #120 - 90 - 60 
      if(adfTest == FALSE){
        pairData$adfTest[end]  <-  1 
      }
      else {
        if(adf.test(pairData$spread[(i-120):end], k = 1)[1] <= criticalValue){
          if(adf.test(pairData$spread[(i-90):end], k = 1)[1] <= criticalValue){
            if(adf.test(pairData$spread[(i-60):end], k = 1)[1] <= criticalValue){
              #If co-integrated then set the ADFTest value to true / 1
              pairData$adfTest[end]  <-  1           
            }
          }
        }
      }
      
      #Calculate the remainder variables needed
      if(i >= mean){
        #Generate Row values
        pairData <- GenerateRowValue(begin, end, pairData)
        
        #Generate the Signals
        pairData <- GenerateSignal(i, pairData)
        
        currentSignal  <-  pairData$signal[i]
        prevSignal  <-  pairData$signal[i-1]
        
        #Generate Transactions
        pairData <- GenerateTransactions(currentSignal, prevSignal, i, pairData)
        
        #Get the returns with added slippage
        pairData <- GetReturns(i, pairData, slippage)
      }
    }
  }
  
  if(generateReport == TRUE)
    GenerateReport(pairData, startDate, endDate)
  
  return(pairData)
}

```

###BacktestPortfolio
BacktestPortfolio accepts a vector of csv files and then generates an equaly weighted portfolio.

Functions arguments: 

* names = an attomic vector of csv file names, example: c('DsyLib.csv', 'OldSanlam.csv')
* mean = the number of observations used to calculate the mean of the spread.
* leverage = how much leverage you want to apply to the portfolio

```{r}
#An equally weighted portfolio of shares
BacktestPortfolio  <- function(names, mean = 35,leverage = 1, startDate = '2005-01-01', endDate = '2015-11-23'){
  ##Itterates through all the pairs and backtests each one
  ##stores the data in a list of numerical vectors
  returns.list  <- list()
  counter  <-  F
  ticker  <- 1
  for (name in names){
    #A notification to let you know how far it is
    #print(paste(ticker, " of ", length(names)))
    #ticker  <- ticker + 1
    
    #Run the backtest on the pair
    data <- read.csv(name)   
    BackTest.df <- BacktestPair(data, mean, generateReport = FALSE)
    
    #Store the dates in a seperate vector
    if (counter == F){
      dates  <<- as.Date(BackTest.df$Date)
      counter  <- T
    }
    
    #Append to list
    returns.list  <- c(returns.list, list(BackTest.df[,18]))
  }
  
  ##Aggregates the returns for each day and then calculates the average for each day
  total.returns  <- c()
  for (i in 1:length(returns.list)){
    if(i == 1)
      total.returns = returns.list[[i]]
    else
      total.returns = total.returns + returns.list[[i]]
  }
  
  total.returns <- total.returns / length(returns.list)
  
  ##Generate a report for the portfolio
  returns  <-  xts(total.returns * leverage, dates)
  GenerateReport.xts(returns, startDate, endDate)
  
  return(returns)
}
```



#Running Backtests
Now we can start testing strategies using our code. 

##Pure arbitrage on the JSE
WHen starting this project the main focus was on using statistical arbitrage to find pairs that were co-integrated and then to trade those, however I very quickly realised that the same code could be used to trade shares that had both its primary listing as well as access to its secondary listing on the same exchange. 

If both listings are found on the same exchange, it opens the door for a pure arbitrage strategy due to both listings refering to the same asset. Therefore you dont need to test for co-integration.

There are two very obvious examples on the JSE. 

###First Example Investec:
Primary = Investec Ltd : Secondary = Investec PLC

####Investec In-Sample Test (2005-01-01 - 2012-11-23)
Test the following parameters

* The Investec ltd / plc pair
* mean = 35
* Set adfTest = F (Dont test for co-integration)
* Leverage of x3

```{r}
#Investec
leverage <- 3
data <- read.csv('investec.csv') 
investec <- BacktestPair(data, 35, generateReport = F, adfTest = F) 

#Format to an xts object and pass to GenerateReport.xts()
investec.returns  <-  xts(investec[,18] * leverage, investec$Date)
GenerateReport.xts(investec.returns, startDate = '2005-01-01', endDate = '2012-11-23')
```

####Investec Out-of-Sample Test (2012-11-23 - 2015-11-23)
Note: if you increase the slippage, you will very quickly kiss profits good bye.
```{r}
GenerateReport.xts(investec.returns, startDate = '2012-11-23', endDate = '2015-11-23')
```


###Second Example Mondi:
Primary = Mondi Ltd : Secondary = Mondi PLC

####Mondi In-Sample Test (2008-01-01 - 2012-11-23)
Test the following parameters

* The Mondi ltd / plc pair
* mean = 35
* Set adfTest = F (Dont test for co-integration)
* Leverage of x3

```{r}
data <- read.csv('mondi.csv') 
mondi <- BacktestPair(data, 35, generateReport = F, adfTest = F)

mondi.returns  <-  xts(mondi[,18] * leverage, mondi$Date)
GenerateReport.xts(mondi.returns, startDate = '2008-01-01', endDate = '2012-11-23')
```

####Mondi Out-of-Sample Test (2012-11-23 - 2015-11-23)
Note: In all of my testing I found that the further down the timeline my data was, the harder it was to make profits on end of day data. I tested this same strategy on intradaydata and it has a higher return profile.

```{r}
GenerateReport.xts(mondi.returns, startDate = '2012-11-23', endDate = '2015-11-23')
```


#Statistical Arbitrage on the JSE
Next we will look at a pair trading strategy.

Typically a pair consits of 2 shares that:

* Share a market sector
* Have a similar market cap
* Similar business model and clients
* Are co-integrated

In all of the portfolios below I use 3x leverage

##Contruction Portfolio
###In-sample test (2005-01-01 - 2012-11-01)
```{r, warning=FALSE}
names  <- c('groupmr.csv', 'groupppc.csv', 'groupavenge.csv', 'groupwhbo.csv', 
            'mrppc.csv', 'mravenge.csv')

ReturnSeries  <- BacktestPortfolio(names, startDate = '2005-01-01', endDate = '2012-11-01', leverage = 3)
```

###Out-of-sample test (2012-11-23 - 2015-11-23)
```{r}
GenerateReport.xts(ReturnSeries, startDate = '2012-11-23', endDate = '2015-11-23')
```


##Insurance Portfolio
###In-sample test (2005-01-01 - 2012-11-01)
```{r, warning=FALSE}
names  <- c('disclib.csv', 'discmmi.csv', 'discsanlam.csv', 'libmmi.csv', 'mmiold.csv',
            'mmisanlam.csv', 'oldsanlam.csv')

ReturnSeries  <- BacktestPortfolio(names, startDate = '2005-01-01', endDate = '2012-11-01', leverage = 3)
```
###Out-of-sample test (2012-11-23 - 2015-11-23)
```{r}
GenerateReport.xts(ReturnSeries, startDate = '2012-11-23', endDate = '2015-11-23')
```


##General Retail Portfolio
###In-sample test (2005-01-01 - 2012-11-01)
```{r, warning=FALSE}
names  <- c('wooltru.csv', 'woolmr.csv', 'wooltfg.csv', 'trumr.csv', 'trutfg.csv', 'MRTFG.csv')

ReturnSeries  <- BacktestPortfolio(names, startDate = '2005-01-01', endDate = '2012-11-01', leverage = 3)
```

###Out-of-sample test (2012-11-23 - 2015-11-23)
```{r}
GenerateReport.xts(ReturnSeries, startDate = '2012-11-23', endDate = '2015-11-23')
```


#Notes:

* From the above sector portfolios we can see that the early years are very profitable but the further down the timeline we go, the lower returns get. 
* The same is noticed with other portfolios that I didnt include in this report but is in the R Code file.
* I believe that this is due to large institutions using the same bread and butter strategy. You will note (if you spend enough time testing all the strategies) that in 2009 there seems to be a sudden shift in the data to lower returns.
* I feel that the end of day data I am using is limiting me and if I were to test the strategy on intraday data then profits would be higher. (I ran one test on intraday data on mondi and the results were much higher, but I am still to test it on sector portfolios)
* This is one of the more simple statistical arbitrage strategies and I believe that if we were to improve on the way we calculate the spread and change some of the entry and exit rules, the strategy would become a more profitable.



